Modeling contributions to the IT-related emissions of a university

### 2025-06-22 doc/paper.tex
Skeleton for paper submission

\addbibresource{ref.bib}
\title{UZH IT \& AI Emissions Analysis\vspace{-1em}}
\author{Massimiliano Poletto}
\date{\vspace{-1em}Last modified: 18.12.2024}

@@

\maketitle
\section{Introduction}

We are trying to answer two questions:
\begin{itemize}
    \item Are there emissions savings in moving UZH's IT infrastructure to the cloud?
    \item What is the impact of increased AI use (and, in particular, of LLMs) on UZH's emissions?
\end{itemize}

To this end, we need some UZH-specific information:
\begin{enumerate}
    \item The size of UZH's student and staff population and a rough breakdown of
        what kind of work they do (to estimate their use of IT and AI).
    \item Information about UZH's current IT infrastructure (number of servers,
        classroom equipment, etc.) and the energy use (electricity bills) of any
        campus data centers.
    \item An estimate of the cloud compute capacity required to host UZH's current IT requirements.
\end{enumerate}

In addition, we need information about the possible location of cloud data centers and
LLM deployments, the embodied emissions of different types of IT equipment, and the carbon intensity
of electricity in different regions.

We describe our modeling approach and our assumptions and uncertainties regarding these factors.

\section{Modeling approach}

At a high level, we model IT emissions primarily bottom-up, based on device counts and
estimates of their characteristics (embodied emissions, power utilization, and duty cycle).

The model can operate in two modes: deterministic, where each device type has a fixed set of
characteristics based on the mean of values we found in the literature, and stochastic, where
we pick device characteristics from truncated normal distributions with the mean and standard
deviation of the literature values. Of course, normality is a strong and probably incorrect
assumption, but it is a simple initial way to model the uncertainty in our estimates.

\subsection{Campus emissions}

We instantiate as many devices as specified in the UZH spreadsheet, plus enough laptops and phones
to cover the rest of the university population.

For each device instance, we pick embodied emissions, power draw, and duty cycles either
deterministically or by sampling from the appropriate distribution.

Combining power draw and duty cycle estimates with ground-truth data about the Swiss electricity mix,
we estimate annual operational emissions and add them to the embodied emissions scaled by the device
type's expected lifetime.

For on-campus data centers, for which presumably there are electricity bills or other measures of
overall energy use, we would like to use that data to evaluate our model.

A possible future refinement will be to model the duty cycle as a function of time of day
and incorporate finer-grained electricity mix data.

\subsection{Cloud emissions}

We model cloud emissions based on the assumptions and data described in
Section~\ref{sec:cloud_data_centers}. We suggest working with UZH's cloud vendor to obtain
more accurate data.

\subsection{LLM emissions}

We model LLM emissions by sampling from the per-capita daily query distribution and computing
the carbon intensity of each query based on the average European electricity mix. While there 
are many uncertainties about the level of LLM adoption, there appears to be reasonable consensus
in the literature about the energy consumption of each query.

\section{Data sources and unknowns}

\subsection{UZH population}

We have access to a spreadsheet (\texttt{2024-11-13\_Berechnung\_THG\_IKT\_2024.xls}) with
counts and emissions factors for various types of IT equipment at UZH. It reports approximately
10,000 laptops and monitors, 1,000 servers, 2,400 mobile phones,  150 classroom PCs, and a few
hundred printers, copiers, and projectors.

By contrast, \textcite{uzhprofile} reports approximately 28,000 students and 7,300 staff
at UZH. Evidently the spreadsheet only lists university assets, not personal devices.

\assumption{In modeling overall university IT emissions as well as potential LLM use,
we assume that 90\% of the university population, 32,000 people, have a laptop and a mobile phone.}
We do not yet have any information about the breakdown of different functions / job roles, so
\assumption{we assume that IT use is uniform across the university}. Both of these assumptions
should be revisited.

\subsection{Embodied emissions}

For several device types, especially end-user devices such as phones and laptops, embodied emissions
comprise a majority of lifetime emissions. However, the exact amount of these emissions varies
widely between different sources. Our approach has been to find several studies and industry references,
and then estimate means and standard deviations from those that seemed most comprehensive and reliable
(for example, Dell's Product Carbon Footprint reports about its servers).

\subsubsection{Laptops}

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
121 & \textcite{ecoinvent}  \\ \hline
104 & \textcite{teehan2013} (low) \\ \hline
338 & \textcite{teehan2013} (high) \\ \hline
244 & \textcite{rarecoil} \\ \hline
202 & \textcite{unctadder2024} \\ \hline
\end{tabular}
\label{tab:embodied_emissions:laptops}
\end{center}

The Ecoinvent numbers appear to be outliers on the low end.
The most comprehensive data comes from \textcite{rarecoil}, which lists manufacturer
information for over 90 popular devices. Mean supply chain emissions are
244 kg CO$_2$e with a standard deviation of 128 kg.
\assumption{We model laptop supply chain emissions using a truncated normal distribution with those characteristics.}

\subsubsection{Desktops}

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
238 & \textcite{ecoinvent} \\ \hline
303 & \textcite{teehan2013} \\ \hline
403 & \textcite{unctadder2024} \\ \hline
289 & \textcite{dellpcf} \\ \hline
277 & \textcite{boavizta} \\ \hline
\end{tabular}
\label{tab:embodied_emissions:desktops}
\end{center}

Most estimates are in the range of 250 - 400 kg CO$_2$e. The most detailed data comes from
Dell's Product Carbon Footprints (\textcite{dellpcf}), which they publish for each of their products.
A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
\assumption{We model desktop supply chain emissions using these statistics.}

\subsubsection{Servers}

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
383 & \textcite{teehan2013} \\ \hline
1252 & \textcite{davy2021} (Dell) \\ \hline
1912 & \textcite{davy2021} (EC2) \\ \hline
899 & \textcite{boavizta} (medium server)\\ \hline
\end{tabular}
\label{tab:embodied_emissions:servers}
\end{center}

\textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
The former have mean supply chain emissions of 1252 kg CO$_2$e with $\sigma = 330$ kg, the latter
1912 kg with $\sigma = 885$ kg.
\assumption{We model server supply chain emissions using the statistics corresponding to Dell servers.}

\subsubsection{Smartphones}

\begin{center}
\begin{tabular}{|r|l|}
\hline
\textbf{Production \& Disposal} & \\
\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
50 & \textcite{unctadder2024} \\ \hline
50 & \textcite{lovehagen2023} \\ \hline
\end{tabular}
\label{tab:embodied_emissions:phones}
\end{center}

\assumption{We model phone supply chain emissions using mean 50 kg CO$_2$e and $\sigma = 10$ kg.}

\subsubsection{Other devices}

We found less data for other types of devices. For now, we model them with averages based
on a handful of sources.

\begin{description}
    \item[Computer monitors] 344 kg CO$_2$e (\textcite{teehan2013}, \textcite{dellpcf})
    \item[Conference room displays] 753 kg CO$_2$e (scaling monitor by $(40''/27'')^2$)
    \item[Printer/copier stations] 1167 kg CO$_2$e (\textcite{ecoinvent})
    \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
      and have no information about UZH network architecture. The vast majority ($80-95\%$) of
      lifecycle emissions of network gear stem from usage (\textcite{cisco2024}, \textcite{jacob2023}).
      For now, we model network equipment as a 5\% overhead on servers (1 router or switch for 20 servers).
      (A typical data center rack contains 42 1U servers and one switch or router.)
\end{description}

\subsection{Operational emissions}

For on-premise equipment, we would like to model the Swiss electricity mix (see below) and, ideally,
daily variations in usage and electricity carbon intensity.
Generic manufacturer estimates of lifetime usage-related emissions are therefore unsuitable.
There are surprisingly few sources of data on device power consumption (as opposed to CO$_2$ emissions).
Based on a smattering of data sheets, primarily from Apple, Cisco, and Dell, we model
power draw of different devices while under load using the following initial parameters:

\begin{center}
\begin{tabular}{|l|r|r|}
\hline
\textbf{Device} & \textbf{Mean (W)} & \textbf{$\sigma$ (W)} \\ \hline
Laptop & 30 & 5 \\ \hline
Desktop & 100 & 20 \\ \hline
Server & 400 & 100 \\ \hline
Phone & 5 & 2 \\ \hline
Monitor & 50 & 10 \\ \hline
Conference display & 250 & 50 \\ \hline
Printer/copier & 1000 & 200 \\ \hline
\end{tabular}
\end{center}

We assume a distribution of duty cycles and power draw when idle for each device.
In the future, we may simulate power draw at different times of day to integrate with
hourly electricity mix data, but errors in those assumptions may swamp the relatively small
hourly variations in Swiss electricity carbon intensity.

\subsection{Cloud data centers}
\label{sec:cloud_data_centers}

The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud
infrastructure. The best resource we have been able to find is the Datavizta API (\textcite{boavizta}).

Lacking information about UZH's workloads, we model a hypothetical UZH cloud footprint as follows:
\begin{itemize}
    \item One-for-one replacement of one campus server with one Azure D8S\_v3 instance (2x Intel
      Xeon, 8 cores, 32 GB RAM, 2TB SSD) (\cite{msftvms}).
    \item Deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region .
    \item Estimation of annual emissions via Datavizta, assuming constant 50\% load: 31 kg CO$_2$e for usage,
      97 kg CO$_2$e for manufacturing.
\end{itemize}

\subsection{Large language models}

Energy consumption of LLMs is being studied extensively (\textcite{budennyy2022eco2ai},
\textcite{castano2023exploring},
\textcite{devries2023growing},
\textcite{gowda2024watt},
\textcite{harding2024watts}, 
\textcite{heguerte2023estimate},
\textcite{luccioni2022estimating},
\textcite{luccioni2023counting},
\textcite{patterson2021carbon}, 
\textcite{rodriguez2024evaluating},
\textcite{tripp2024measuring}).

The best estimates are that use of large language model (LLM) like ChatGPT averages 3-4 Wh / request,
or approximately 10x the energy of a traditional search engine query. This number appears to be decreasing
rapidly as model hardware and software improve, even though overall energy use is increasing due to
increased utilization. Nevertheless, for the model we assume that queries consume 3 Wh each.
We assume that on average every user will issue 15 queries per day.

LLMs are distributed and the location that serves any particular request is unknown, so we use the 
average European electricity mix to compute carbon intensity.

The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely.
However, \textcite{devries2023growing} provides the following information about ChatGPT:
\begin{itemize}
    \item Operational power consumption of 500 MWh / day.
    \item Approximately 1300 MWh consumed in training.
    \item Approximately 4000 servers.
\end{itemize}

Assuming a relatively clean grid at 200 g CO$_2$e / kWh (a low estimate), and that each server has embodied emissions of 4 t CO$_2$e (a high estimate),
we obtain:
\begin{itemize}
    \item 100 t CO$_2$e / day operational emissions.
    \item 260 t CO$_2$e training emissions.
    \item 16 kt CO$_2$e total embodied emissions.
\end{itemize}

After one year of operation, operational emissions are approximately 36 kt CO$_2$e, whereas training and embodied emissions are about 16 kt CO$_2$e,
so operational emissions account for $\approx 70\%$ of the total. With less conservative assumptions (dirtier grid, lower embodied emissions),
the proportion of operational emissions increases further.

As a result, when modeling overall LLM emissions, we multiply estimated operational emissions by 1.4 to account for training and embodied emissions.
We acknowledge that this is, at best, a rough estimate based on just one LLM.

\subsection{Electricity carbon intensity}

The Swiss Federal Office for the Environment published an extensive report (\textcite{krebs2018umweltbilanz})
on the environmental impact of electricity generation in Switzerland in 2018.

However, for the purposes of the model, we use data from \textcite{electricitymaps}. Specifically,
we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023
at hourly granularity, then aggregate the data into two-hour blocks, and compute the mean of each two
hour block over the course of the year. This gives us twelve data points that describe average
carbon intensity variation over the course of the day.

\section{Missing data}

The following data would improve the accuracy of our model (in roughly decreasing order of utility):
\begin{itemize}
    \item Campus data center energy use (e.g., at the granularity of monthly electricity bills).
    \item Current campus data center workloads (to better model transition to the cloud).
    \item Academic calendar (e.g., how many days per year is the university full vs operating
        at lower levels).
    \item Common or otherwise representative pieces of equipment (servers, routers, standard desktops, etc.).
    \item Rough breakdown of UZH's population by field / role.
\end{itemize}

\section{Other considerations}

\subsection{Other IT components}

For now, we do not attempt to model changes in network infrastructure or utilization due to
move to the cloud or increased LLM use. See \textcite{jacob2023} for an overview of
network sustainability.

\subsection{Potential AI benefits}

We typically think of AI use as creating additional energy demands and GHG emissions.
So far, this model does not include potential energy savings caused by AI, such as improvements
to building energy management, campus logistics, etc.

\textcite{tomlinson2024carbon} make the argument that tasks such as writing
and illustrating can be done with fewer GHG emissions by AI than by humans. They compare the
estimated emissions of an LLM performing the task to those of the average human, calculated
by multiplying per-capita annual emissions by the time ($\approx 1$h) required to write or draw.
Using this approach, AI is orders of magnitude cheaper than humans, whether one uses per-capita
emissions from the US (15 t CO$_2$e/y) or India (1.9 t). Of course, this methodology focuses on
a specific creative act: the LLM does not reduce the human's overall annual emissions.

\printbibliography
\end{document}

>>>>

\addbibresource{ref.bib}
\title{Modeling contributions to the IT-related emissions of a university}
\author{Massimiliano Poletto}

@@

\maketitle

\section{Abstract}

\section{Introduction}

\section{Modeling methodology}

\section{Results}

\section{Related work}

\section{Conclusion}

\section{References}

\end{document}
